Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
section 4 edits

\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus our model validates the specification synthesis process which concludes the
inference procedure. For our example, we infer and then synthesise the
5-point specification:
%%
\begin{minted}[linenos=false]{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
\end{minted}
%%

@@

\begin{example}
We give an extended example of verifying a stencil computation against
its specification. We consider a more general, but still consistent, version of the \textit{five-point}
stencil specification inferred previously in Section~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

Verification of the code against the spec is then given by 
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
We explain the steps.
\begin{enumerate}  
%
\item Generate the model from the specification as in
  Section~\ref{sec:semantics}. 
%
\begin{align*}
 M_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\[0.4em]
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (M_s))
\end{align*}
%
\item Obtained the code model via $\textit{modelise}$, giving:
%
\begin{align*}
M_c = \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\\[0.4em]
\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
\end{align*}
%
\item Consistency $cons_2(\textsf{once}(\textsf{exact}(M_s)),
  \textsf{once}(M_c))$ holds when 
  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(M_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
  which holds when: 
  \begin{align*}
     & \textit{cons}_2(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c)) \\
   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(M_s),\textsf{mult}(M_c))
  \end{align*}
%
\item This condition holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in M_s . u \preceq_2 v$: 
  \begin{align*}
    \begin{array}{l}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \end{align*}
  % 
  and for the lower bound: $\forall v \in M_s, \exists u \in M_c . u
  \preceq_n v$: 
  \begin{align*}
    \begin{array}{l}
     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
  \end{align*}

  Since both the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{enumerate}
\noindent
Now when $L3$ is modified with an additional term \texttt{b(i+2,j+2)},
the new assignment $L3'$ becomes:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
In the first six steps, evaluation proceeds the same way except for step three.
In the third step, result of $\mathit{modelise}(L3',b)$ will have the extra term
$\vtwoh{2}{2}$. In the seventh step, since the model has not changed and the
indexing schemes are only augmented with a new element, the generalisations for
lower bound still hold. However, there are no elements in the model, $m$, that
can generalise $\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.

Note that if the specification were as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
The consistency predicate would have held because for the upper bound
$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m'))$ would not have
been needed. This semantically makes sense since the region $\texttt{fivepoint}$
is indeed contained in the regions touched by $b$.
\end{example}

As noted above, the \textsf{fivepoint} region defined above is more accepting
than that used in Section~\ref{sec:lang}. This introduces a new notion of
approximation beyond that is achieved in Section~\ref{sec:semantics}
using upper and lower bounds because the in both cases the specifications are
used to generate \emph{exact} models. Formally, we say a specification $s'$
approximates a more specific specification $s$ \emph{iff} $\forall m \in
\textsf{Mult}(\domainVal) . \mathit{cons}(s,m) \implies \mathit{cons}(s',m)$.

Equations provided on regions in Section~\ref{sec:lang} when all modifiers
equal provide such approximations, but it goes further than that. The
regions describing the fivepoint stencil for example cannot be shown equal
using these equations, but satisfies the approximation predicate as defined
above.

The checking algorithm though happily accepts an approximation in place of the
most specific specification. Although this decreases the likelihood of catching
stencil errors (as more access patterns conform), it reduces the space and
effort it takes to write the specification.

>>>>

\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus our model validates the specification synthesis process which concludes the
inference procedure. For our example, we infer and then synthesise the
5-point specification:
%%
\begin{minted}[linenos=false]{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1):: a
\end{minted}
%%

@@

\begin{example}
We give an extended example of verifying a stencil computation against
its specification. We consider a more general, but still consistent, version of the \textit{five-point}
stencil specification inferred previously in
Section~\ref{subsec:inference} (see top-left):
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

Verification of the code against the spec is then given by 
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
We explain the steps.
\begin{enumerate}  
%
\item Generate the model from the specification as in
  Section~\ref{sec:semantics}. 
%
\begin{align*}
 M_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\[0.4em]
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (M_s))
\end{align*}
%
\item Obtained the code model via $\textit{modelise}$, giving:
%
\begin{align*}
M_c = \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\\[0.4em]
\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
\end{align*}
%
\item Consistency $cons_2(\textsf{once}(\textsf{exact}(M_s)),
  \textsf{once}(M_c))$ holds when 
  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(M_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
  which holds when: 
  \begin{align*}
     & \textit{cons}_2(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c)) \\
   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(M_s),\textsf{mult}(M_c))
  \end{align*}
%
\item This condition holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in M_s . u \preceq_2 v$: 
  \begin{align}
    \begin{array}{l}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \label{eq:check-example-upper}
  \end{align}
  % 
  and for the lower bound: $\forall v \in M_s, \exists u \in M_c . u
  \preceq_n v$: 
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
    \label{eq:check-example-lower}
  \end{align}
  Since both the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{enumerate}
\vspace{-0.7em}
\end{example}
\begin{example} 
We consider now an erroneous user modification to $l3$ with an
additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The checking procedure proceeds in the same way as the first three
steps of the previous example, except that the result of
$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
$\vtwoh{2}{2}$. Since the spec model has not changed and the code
model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
the lower bound \eqref{eq:check-example-lower} 
of step 4 above still holds since its consistency condition
universally quanitifes over the \emph{spec model}. However, for
the upper bound, the consistency condition universally quantifies
over scheme $u$ of the \emph{code model} for which there exists
a scheme $v \in M_s$ such that $u \preceq_2 v$. Thus, we require a $v
\in M_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
scheme existing in the spec model. Hence the model fails to hold as an upper bound.

Note that if the specification were as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
then consistency holds as it no longer requires the failing upper-bound case 
$\mathit{cons}(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c))$.
 This makes sense semantically since the region ${\textcolor{darkgreen}{\texttt{fivepoint}}}$
is indeed contained in the regions touched by $\texttt{b}$ in $l3'$. 
\end{example}

As noted, the {\textcolor{darkgreen}{\term{fivepoint}}} region
 above is more accepting than that used in
Section~\ref{sec:lang}. This introduces a new notion of approximation
beyond what is given by the syntactic theory of sub-specifications
$<:$ in Section~\ref{sec:eqs}. Consistency leads to a notion of
approximation that is a \textit{semantic sub-specification} relation:
%using upper and lower bounds because the in both cases the specifications are
%used to generate \emph{exact} models.

\begin{definition}[Approximation]
A specification $S'$
\emph{approximates} a more specific specification $S$, written 
$S \sqsubseteq S'$, \emph{iff}:
$$
\forall M_c \in \textsf{Mult}(\domainVal{}^n) . \mathit{cons}_n(\interp{S}_n,M_c) \implies
\mathit{cons}(\interp{S'}_n,M_c)
$$
\end{definition}
%
For the syntactic sub-specification equations on regions $<:^r$, it follows that
these imply semantic approximation:
%
\begin{lemma}[Sub-specification $<:^r$ coincides with approx 
$\sqsubseteq$]
\begin{equation*}
\forall S, S' . \; S <:^r S' \Rightarrow S \sqsubseteq S'
\end{equation*}
\end{lemma}
%
However, the converse direction does not hold. 
The regions describing the more acceptiong five-point stencil
above do not follow from the definition of $<:$ on the previous
more specific specification.%, that is:
%${\textcolor{darkgreen}{\term{centered(depth=1, dim=1)*reflexive(dim=2)
%      + centered(depth=1, dim=2)*reflexive(dim=1)}}}
%\not{<:} {\textcolor{darkgreen}{\term{centered(depth=1,dim=1) +
%      centered(depth=1,dim=2)}}}$.
 However, consitency (and thus $\sqsubseteq$) shows
that we can check the five-point stencil code against the more general
specifcation of {\textcolor{darkgreen}{\term{fivepoint}}}. Thus, we
can always use an approximation in place of the
more specific specification. Although this decreases the likelihood of catching
stencil errors (as more access patterns conform), it reduces the space and
effort it takes to write the specification.
